A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-15 03-semantics.tex
Reformatação do inicio da Seção 3.2
Comprimi a definição dos conjuntos sintáticos (P, N e E).

The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system calls
that affect a memory location identified by~$id$.
%
According to the synchronous hypothesis of \CEU, $\ceu{\Mem}$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
share identifiers, i.e., an identifier is either a variable, an external event,
or an internal event.

Regarding mismatches between the concrete and abstract languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
\code{v=10}.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an \code{emit} and access it after the
corresponding \code{await}.
%
Also, a ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
In the concrete language, \code{<A>} and \code{<C>} execute in sequence, and
the finalization code \code{<B>} is implicitly suspended waiting for \code{<C>}
termination.
In the formal language, ``\code{fin <B>}'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
termination aborts ``\code{fin <B>}'', which finally causes \code{<B>} to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program $p$, a stack level $n$, and an
emitted event $e$ to a modified triple as follows:
%
\begin{align*}
\LL p,  n,  e  \RR &\1\xrightarrow\1
\LL p', n', e' \RR
    & \textbf{(rule-general)}
\end{align*}
%
where
%
\begin{align*}
p, p' &\in P
    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
\\
n, n' &\in \mathbf{N}
    && (current~stack~level)
\\
e, e' &\in \mathbf{E} \cup \{\epsilon\}
    && (emitted~event,~possibly~none)
\end{align*}
%
At the beginning of a reaction to an input event $id$, the triple is
initialized at stack level $0$ ($n=0$) and with the emitted event $id$
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the triple will block with a (possibly) modified program $p'$, at stack level
$0$, with no event emitted $\epsilon$:
%
\begin{align*}
\LL p,0,ext \RR
    & \1\xrightarrow{*}\1
\LL p',0,\epsilon \RR
\end{align*}
%
We also distinguish between two transition rule subtypes:
    \emph{outermost} transitions $\OUT$; and
    \emph{nested} transitions $\NST$.

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level as
follows:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level, and, at the same time,
    (c) consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see $isblocked(n,p)$ in
Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.

>>>>

The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system calls
that affect a memory location identified by~$id$.
%
According to the synchronous hypothesis of \CEU, $\ceu{\Mem}$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).

We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
share identifiers: any identifier is either a variable, an external event,
or an internal event.

Most expressions in the abstract language are mapped to
their counterparts in the concrete language.  The exceptions are the
finalization block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions, which are
internal expressions that result from the
expansion of awaits, emits, and loops by the transition rules to be discussed.

Regarding other mismatches between the concrete and abstract languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
corresponding $\ceu{\AwaitInt}$.
%
Also, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete language is
equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract language.
In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
termination.
In the abstract language, ``$\ceu{\Fin B}$'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute~$C$ in parallel, whose
termination aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p, n, e>\to\<p',n',e'>
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
nonnegative integers representing the current stack level,
and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after
the transition, both possibly the empty event~($\nil$).

% \begin{align*}
% p, p' &\in\P
%     && (program~as~described~in~Listing~\ref{lst.formal.syntax})
% \\
% n, n' &\in\N
%     && (current~stack~level)
% \\
% e, e' &\in\E \cup \{\epsilon\}
%     && (emitted~event,~possibly~none)
% \end{align*}
%
At the beginning of a reaction to an input event~$id$, the triple is
initialized at stack level $0$ ($n=0$) and with the emitted event
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the triple will block with a (possibly) modified program~$p'$, at stack
level~$0$, and with no event emitted~$(\nil)$:
%
\[
  \<p,0,e>\xrightarrow{*}\<p',0,\nil>
\]

We distinguish between two types of transition rules:
    \emph{outermost} transitions $\OUT$ and
    \emph{nested} transitions $\NST$.

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level, and, at the same time,
    (c) consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see $isblocked(n,p)$ in
Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.
